---
title: "Macroinvertebrate Data Analysis"
author: "Your Name Here"
date: "10/__/2020"
output: html_document
---

# Load packages


```{r load packages, echo=TRUE, message=FALSE, warning=FALSE}

library(tidyverse) ## for readr, dplyr, ggplot
library(lubridate) ## for dealing with dates and times
library(vegan) ## for calculating diversity measures
library(ggbeeswarm) ## for jittering points

```


# Import Prepared Data

There are several files in the "preppedData" folder.  Below is the code to import each file.  Delete any code chunks you don't need (only import the data you actually need)


## MILL Data only

```{r import MILL only, include=FALSE}

# this chunk will import the data for the Mill River only
# this file includes all the Mill River data including all water quality and weather data
# see the file called mill_macro_metadata.txt for details about all these variables and methods

mill.macro <- read_csv("preppedData/mill.macro.csv",
          col_types = cols(date = col_date(format = "%m/%d/%Y"),
                           sampleTime = col_factor(levels=c("Summer 2018", 
                                                            "Fall 2018",
                                                            "Summer 2019", 
                                                            "Fall 2019")),
                          location = col_factor(levels = c("Downstream",
                                                           "Upstream")), 
                          season = col_factor(levels = c("Summer", 
                                                         "Fall")), 
                          year = col_factor(),
                          microhabitat = col_factor(levels = c("DSR", 
                                                               "DFR", 
                                                               "DM", 
                                                               "DSH", 
                                                               "DSP",
                                                               "USR", 
                                                               "UFR", 
                                                               "UM", 
                                                               "USH", 
                                                               "USU")))) %>% 
      mutate(siteID = "MILL")




```

## All NEON Macroinvertebrate Data

```{r import all NEON macro data, include=FALSE}

# this chunk will import data for ALL NEON macroinvertebrate data for all wadeable streams
# this dataset does not have any environmental variables added

neon.macro <- read_csv("NEONData/neon.macro.csv") %>% 

        # add date and year to data
          dplyr::mutate(date = date(collectDate),
                        year = year(collectDate))



```


## Master Taxa List

```{r import master taxa list, include=FALSE}

#this will import the list of all taxa for the Mill River and NEON rivers

master.taxa <- read_csv("data/neon.mill.taxa.csv")


```



## Notes about NEON Aquatic Instrument Data

The Aquatic Instrument Data come from sensors placed upstream and downstream of the sampling reach. See the section on Placement of Sensors by Site Type for Wadeable Streams on the [Surface Water webpage](https://www.neonscience.org/data-collection/surface-water).  Please note that these are not equivalent to the Upstream and Downstream locations in the Mill River



## HOPB Data only

```{r import HOPB only, include=FALSE}

# this chunk will import the data for the HOPB River only
# this file includes the macroinvertebrates and most of the environmental variables

# this file has the data from the upstream sensor (no fDOM)
HOPB.macro.up <- read_csv("preppedData/HOPB.macro.up.csv")

# this file has the data from the downstream sensor (includes fDOM)
HOPB.macro.down <- read_csv("preppedData/HOPB.macro.down.csv")

```






# Diversity

```{r data wrangling for diversity measures}

# remove missing values, and make sure each taxa is summarized within sampleID
macro.long <- mill.macro %>% 
  dplyr::filter(!is.na(number)) %>% 
  dplyr::select(sampleID, acceptedTaxonID, number) %>% 
  group_by(sampleID, acceptedTaxonID) %>% 
  dplyr::summarise(num = sum(number))


# convert to wide format
macro.wide <- macro.long %>% 
    pivot_wider(names_from = acceptedTaxonID, 
              values_from = num,
              values_fill = list(num = 0),
              values_fn = list(num = mean)) %>%
    tibble::column_to_rownames("sampleID")


```

```{r calculate diversity measures}

#Shannon
H <- diversity(macro.wide)


#effective number of species
effective.sp <- exp(H)


#Richness
sp.rich <- specnumber(macro.wide) 

#Max possible diversity
max.H <- log(sp.rich)


#Pielou's Evenness J 
J <- H / max.H

#put all diversity values into a single data frame
macro.div <- data.frame(H, effective.sp, sp.rich, max.H, J)



```

# Tolerance

```{r calculate tolerance}

# remove missing values, and make sure each taxa is summarized within sampleID
macro.tol <- mill.macro %>% 
  dplyr::filter(!is.na(number), !is.na(tolerance)) %>% 
  dplyr::select(location, sampleID, acceptedTaxonID, number, tolerance) %>% 
  group_by(location, sampleID, acceptedTaxonID) %>% 
  dplyr::summarise(num = sum(number), 
                   tol = mean(tolerance)) %>% 
  mutate(x.t = num*tol)

macro.HBI <- macro.tol %>% 
  group_by(location, sampleID) %>% 
  dplyr::summarise(x.t.sum = sum(x.t),
                   total.n = sum(num)) %>% 
  mutate(HBI = x.t.sum/total.n)


```



```{r  calculate EPT}

# number of Es
macro.E <- mill.macro %>% 
  dplyr::filter(order == "Ephemeroptera") %>% 
  group_by(location, sampleID) %>% 
  dplyr::summarise(num.E = sum(number)) 

# number of Ps
macro.P <- mill.macro %>% 
  dplyr::filter(order == "Plecoptera") %>% 
  group_by(location, sampleID) %>% 
  dplyr::summarise(num.P = sum(number)) 

# number of Ts
macro.T <- mill.macro %>% 
  dplyr::filter(order == "Trichoptera") %>% 
  group_by(location, sampleID) %>% 
  dplyr::summarise(num.T = sum(number)) 

# number of net-spinners
macro.H <- mill.macro %>% 
  dplyr::filter(family == "Hydropsychidae") %>% 
  group_by(location, sampleID) %>% 
  dplyr::summarise(num.H = sum(number)) 

# number of all organisms
macro.all <- mill.macro %>% 
  group_by(location, sampleID) %>% 
  dplyr::summarise(num.all = sum(number)) 

#join dataframes
macro.EPT <- left_join(macro.E, macro.P) %>% 
                  left_join(., macro.T) %>% 
                  left_join(., macro.H) %>%
                  left_join(., macro.all)

#Calculate %EPT
macro.EPT <-  macro.EPT %>% 
  mutate(EPT = (num.E + num.P + num.T)*100/num.all) %>% 
  mutate(EPT2 = (num.E + num.P + num.T - num.H)*100/num.all)




```


# Functional Feeding Groups

```{r relative abundance of FFG's}



# number of scrapers
macro.scr <- neon.macro %>% 
  dplyr::filter(FFG == "scr") %>% 
  group_by(date, sampleID, domainID, siteID, habitatType) %>% 
  dplyr::summarise(num.scr = sum(estimatedTotalCount)) 


# number of filtering collectors
macro.cf <- neon.macro %>% 
  dplyr::filter(FFG == "cf") %>% 
  group_by(date, sampleID, domainID, siteID, habitatType) %>% 
  dplyr::summarise(num.cf = sum(estimatedTotalCount)) 


# number of gathering collectors
macro.cg <- neon.macro %>% 
  dplyr::filter(FFG == "cg") %>% 
  group_by(date, sampleID, domainID, siteID, habitatType) %>% 
  dplyr::summarise(num.cg = sum(estimatedTotalCount)) 

# number of shredders
macro.sh <- neon.macro %>% 
  dplyr::filter(FFG == "sh") %>% 
  group_by(date, sampleID, domainID, siteID, habitatType) %>% 
  dplyr::summarise(num.sh = sum(estimatedTotalCount)) 

# number of predators
macro.prd <- neon.macro %>% 
  dplyr::filter(FFG == "prd") %>% 
  group_by(date, sampleID, domainID, siteID, habitatType) %>% 
  dplyr::summarise(num.prd = sum(estimatedTotalCount)) 


# number of all organisms
macro.all <- neon.macro %>% 
  group_by(date, sampleID, domainID, siteID, habitatType) %>% 
  dplyr::summarise(num.all = sum(estimatedTotalCount)) 

#join dataframes
macro.FFG <- left_join(macro.scr, macro.cf) %>% 
                  left_join(., macro.cg) %>% 
                  left_join(., macro.sh) %>% 
                  left_join(., macro.prd) %>% 
                  left_join(., macro.all)

#Calculate FFG ratios
macro.FFG <-  macro.FFG %>% 
  mutate(sub.stability = (num.scr + num.cf)/(num.sh + num.cg),
         pred.ratio = num.prd/(num.scr + num.cf + num.sh + num.cg),
         relab.scr = num.scr/num.all,
         relab.cf = num.cf/num.all,
         relab.cg = num.cg/num.all,
         relab.sh = num.sh/num.all,
         relab.prd = num.prd/num.all)


```

# Correlation


```{r correlation}

#select just the two variables you want to compare
mill.macro %>% 
  dplyr::select(wTemp, DO) %>% 
  
  #remove missing values for each variable
  dplyr::filter(!is.na(wTemp), !is.na(DO)) %>% 
  
  #perform parametric correlation
  cor()


```


# Join MILL and HOPB data

```{r bind together Fall 2019 MILL and HOPB data }

# select fall 2019 data and certain variables for MILL
mill.macro2 <- mill.macro %>% 
  dplyr::filter(season == "Fall", year == "2019") %>% 
  dplyr::select(date, siteID, sampleID, microhabitat, 
                acceptedTaxonID, number, tolerance, FFG) %>% 
  
  # join the sampleID and microhabitat names into one and replace sampleID
  tidyr::unite(col = sampleID, sampleID, microhabitat, sep = ".", remove = TRUE) %>% 
  
  # calculate invertebrate density 
  group_by(date, siteID, sampleID, acceptedTaxonID, FFG) %>% 
  dplyr::mutate(inv_dens = sum(number)/0.5,
                tolerance = mean(tolerance))
  
  
  # select fall 2019 data and certain variables for HOPB
HOPB.macro2 <- HOPB.macro.up %>% 
  dplyr::filter(date == "2019-10-03", habitatType == "riffle") %>% 
  
  # rename column to match MILL column for number
  dplyr::rename( ., number = estimatedTotalCount) %>% 
  
  # select variables of interest
  dplyr::select(date, siteID, sampleID,       
                acceptedTaxonID, number, tolerance, FFG, inv_dens) 
  
  
# bind the MILL and HOPB data together
mill.hopb <- bind_rows(mill.macro2, HOPB.macro2)
 
```



