---
title: "Macro_Analysis_Bio131"
author: "Marney Pratt"
date: "10/14/2020"
output: html_document
---
## Load required packages

```{r load packages, echo=TRUE, message=FALSE, warning=FALSE}

library(tidyverse) ## for readr, dplyr, ggplot
library(lubridate) ## for dealing with dates and times
library(rnoaa) ## For downloading data from NOAA)
library(dataRetrieval) ## download USGS data
library(zoo) ## for dealing with time series


```

## Background Data Preparation and Import

The raw data were all entered into a shared Google Sheet as macroinvertebrates were collected and identified by various Smith College students and Dr. Marney Pratt, then a spreadsheet file was downloaded and quality checked by Marney Pratt. 

See the metadata file "macro_metadata" (https://docs.google.com/document/d/e/2PACX-1vTgT_48p23wqN40yft9WeKVEI8Eh4IjyB94KdsY2IPQgIjAuL0WaC4WR2zbX9D6TZCbXx3X7eghPbbz/pub) for more details on sampling and definitions of all variables in all datasets

### Import Sampling Data
This file contains information from each sample date including weather, river conditions, and other sampling details

```{r import sampling data}

#import data file
sampling <- read_csv("data/sampling.csv", 
    col_types = cols(date = col_date(format = "%m/%d/%Y"), 
        season = col_factor(levels = c("Spring", "Summer", "Fall")),
        location = col_factor(levels = c("Downstream","Upstream")),
        year = col_factor()))

#calculate the total area sampled (q.area)
sampling <- sampling %>% 
  mutate(q.area = quadsTaken*quadSize)

#Remove spring sampling
sampling <- sampling %>% 
  filter(season != "Spring")


```


### Import Quadrat Data
This file contains information taken for each quadrat during sampling including 

```{r import quadrat data}

##Import the quadrat data file
quadrats <- read_csv("data/quadrats.csv", 
    col_types = cols(date = col_date(format = "%m/%d/%Y"), 
        season = col_factor(levels = c("Spring", "Summer", "Fall")),
        location = col_factor(levels = c("Downstream", "Upstream")),
        year = col_factor(),
        time = col_time(format = "%H:%M"),
        microhabitat = col_factor(levels = 
                    c("DSR", "DFR", "DM", "DSH", "DSP", 
                      "USR", "UFR", "UM", "USH", "USU"))))


```

### Import Water Quality Data
Thie file contains water quality information taken during sampling (Sampling = macros) as well as on separate days when just water quality was measured in both the Upstream and Downsream locations on the same day (Sampling = water)

```{r import water quality data}

#import the water quality data file
waterQ <- read_csv("data/waterQ.csv",
  col_types = cols(date = col_date(format = "%m/%d/%Y"), 
                   location = col_factor(levels = c("Downstream","Upstream")), 
                   season = col_factor(levels = c("Summer","Fall")), 
                   year = col_factor(),
                   microhabitat = col_factor(levels = 
                    c("DSR", "DFR", "DM", "DSH", "DSP", 
                      "USR", "UFR", "UM", "USH", "USU"))))

```

### Import Macroinvertebrate Data


```{r Import raw macro data}

#import updated raw.macro-updated.csv file
raw.macro.updated <- read_csv("data/rawMacro.csv",
          col_types = cols(date = col_date(format = "%m/%d/%Y"),sampleTime = 
          col_factor(levels=c("Fall 2015",
                              "Spring 2016", "Summer 2016", "Fall 2016",
                              "Spring 2017", "Summer 2017", "Fall 2017", 
                              "Summer 2018", "Fall 2018",
                              "Summer 2019", "Fall 2019")),
        location = col_factor(levels = c("Downstream","Upstream")), 
        season = col_factor(levels = c("Spring", "Summer", "Fall")), 
        year = col_factor(),
        sampleID = col_factor(),
        microhabitat = col_factor(levels = c("DSR", "DFR", "DM", "DSH", "DSP",
                       "USR", "UFR", "UM", "USH", "USU")),
        subfamily = col_character(),
        tribe = col_character(),
        notes = col_character()))

###Remove crayfish
macro.data <- raw.macro.updated %>% filter(organism_aggr!="crayfish")

###Remove Spring data
macro.data <- macro.data %>% 
  filter(season !="Spring") 

###Remove Piercers
macro.data <- macro.data %>% 
  filter(FFG != "prc")




```




```{r join macro waterQ and quadrat summer 2018}

#filter waterQ
waterQ.S18 <- waterQ %>% 
  filter(year == "2018", sampling == "macros", season == "Summer") %>% 
  dplyr::select(1:9, 16:18)


#select certain columns of quadrat data
quad.S18 <- quadrats %>% 
  filter(year == "2018", season == "Summer") %>% 
  dplyr::select(date, sampleID, microhabitat, time, quadrat, depth, 
         per_sediment, per_rock, per_organic) %>% 
  
  group_by(sampleID, microhabitat) %>% 
  dplyr::summarise(num_quads = n(), 
            depth = mean(depth, na.rm = TRUE), 
            per_sediment = mean(per_sediment, na.rm = TRUE),
            per_rock = mean(per_rock, na.rm = TRUE), 
            per_organic = mean(per_organic, na.rm = TRUE)) 

abiotic.S18 <- left_join (quad.S18, waterQ.S18)
  

##Join macro data and waterQ
macro.abio.S18 <- macro.data %>% 
  filter(year == "2018", season == "Summer") %>% 
  left_join(., abiotic.S18) %>% 
  mutate(benthicArea = num_quads*0.25) %>% 
  mutate(inv.dens = number/benthicArea)
  
```


```{r join macro waterQ and quadrat fall 2018}

#filter waterQ
waterQ.F18 <- waterQ %>% 
  filter(year == "2018", sampling == "macros", season == "Fall") %>% 
  dplyr::select(1:10, 16:18, 21:22) %>% 
  group_by(date, sampleID, sampling, sky, startTime, endTime) %>% 
  dplyr::summarise(num_w = n(), 
            pH = mean(pH, na.rm = TRUE),
            wTemp = mean(wTemp, na.rm = TRUE), 
            DO = mean(DO, na.rm = TRUE),
            turb = mean(turb, na.rm = TRUE),
            cond = mean(cond, na.rm = TRUE),
            ) 

#select certain columns of quadrat data
quad.F18 <- quadrats %>% 
  filter(year == "2018", season == "Fall") %>% 
  dplyr::select(date, sampleID, microhabitat, time, quadrat, depth, 
         per_sediment, per_rock, per_organic) %>% 
  
  group_by(date, sampleID) %>% 
  dplyr::summarise(num_quads = n(), 
            depth = mean(depth, na.rm = TRUE), 
            per_sediment = mean(per_sediment, na.rm = TRUE),
            per_rock = mean(per_rock, na.rm = TRUE), 
            per_organic = mean(per_organic, na.rm = TRUE)) 

abiotic.F18 <- left_join (quad.F18, waterQ.F18) 

  

##Join macro data and waterQ
macro.abio.F18 <- macro.data %>% 
  filter(year == "2018", season == "Fall") %>% 
  left_join(., abiotic.F18) %>% 
  mutate(benthicArea = num_quads*0.25) %>% 
  mutate(inv.dens = number/benthicArea)
  
```


```{r join macro waterQ and quadrat summer 2019}

#filter waterQ
waterQ.S19 <- waterQ %>% 
  filter(year == "2019", sampling == "macros", season == "Summer") 


#select certain columns of quadrat data
quad.S19 <- quadrats %>% 
  filter(year == "2019", season == "Summer") %>% 
  dplyr::select(date, sampleID, microhabitat, time, quadrat, depth, 
         per_sediment, per_rock, per_organic) %>% 
  
  group_by(sampleID, microhabitat) %>% 
  dplyr::summarise(num_quads = n(), 
            depth = mean(depth, na.rm = TRUE), 
            per_sediment = mean(per_sediment, na.rm = TRUE),
            per_rock = mean(per_rock, na.rm = TRUE), 
            per_organic = mean(per_organic, na.rm = TRUE)) 


abiotic.S19 <- left_join (quad.S19, waterQ.S19) %>% 
  dplyr::select(-notes)
  

##Join macro data and waterQ
macro.abio.S19 <- macro.data %>% 
  filter(year == "2019", season == "Summer") %>% 
  left_join(., abiotic.S19) %>% 
  mutate(benthicArea = num_quads*0.25) %>% 
  mutate(inv.dens = number/benthicArea)
  
```


```{r join macro waterQ and quadrat summer and fall 2019}

#filter waterQ
waterQ.19 <- waterQ %>% 
  filter(year == "2019", sampling == "macros") %>% 
  dplyr::select(-depth, -light, -flow, -notes) %>% 
  group_by(date, sampleID, sampling, sky, startTime, endTime, microhabitat) %>% 
  dplyr::summarise(num_w = n(), 
            lat = mean(lat, na.rm = TRUE),  
            long = mean(long, na.rm = TRUE),  
            alt = mean(alt, na.rm = TRUE),  
            pH = mean(pH, na.rm = TRUE),
            wTemp = mean(wTemp, na.rm = TRUE), 
            DO = mean(DO, na.rm = TRUE),
            turb = mean(turb, na.rm = TRUE),
            cond = mean(cond, na.rm = TRUE),
            aTemp = mean(aTemp, na.rm = TRUE)  
            ) 


#select certain columns of quadrat data
quad.19 <- quadrats %>% 
  filter(year == "2019") %>% 
  dplyr::select(date, sampleID, microhabitat, time, quadrat, depth, 
         per_sediment, per_rock, per_organic) %>% 
  
  group_by(sampleID, microhabitat) %>% 
  dplyr::summarise(num_quads = n(), 
            depth = mean(depth, na.rm = TRUE), 
            per_sediment = mean(per_sediment, na.rm = TRUE),
            per_rock = mean(per_rock, na.rm = TRUE), 
            per_organic = mean(per_organic, na.rm = TRUE)) 

abiotic.19 <- left_join (quad.19, waterQ.19)
  

##Join macro data and waterQ
macro.abio.19 <- macro.data %>% 
  filter(year == "2019") %>% 
  left_join(., abiotic.19) %>% 
  mutate(benthicArea = num_quads*0.25) %>% 
  mutate(inv.dens = number/benthicArea)
  
```



```{r bind all data together}


macro.abio <- bind_rows(macro.abio.19,
                        macro.abio.F18, macro.abio.S18) %>% 
  dplyr::select(-BvA, -sampling, -quadrat, -vial, -aTemp, -num_w)


```


# Import Weather Data for Amherst NOAA station

See this webpage for more information and to cite the data source:
[Global Historical Climatology Network (GHCN)]
(https://www.ncdc.noaa.gov/data-access/land-based-station-data/land-based-datasets/global-historical-climatology-network-ghcn)



```{r Download Daily Values Amherst, eval=FALSE, include=FALSE}

###I ran this code once and saved the output, then I turned off running this chunk

##I used AMHERST: USC00190120
## this imports all the data for the station specified
daily.weather.a <- ghcnd(stationid = "USC00190120", refresh = TRUE)


##this imports the specificed data for the specific date range
daily.weather.tidy.a <- meteo_tidy_ghcnd(stationid = "USC00190120", keep_flags=FALSE, 
                                       var = c("PRCP", "TMAX", "TMIN"),
                                       date_min = "2018-01-01", 
                                       date_max = "2019-12-31")


daily.weather.tidy.a <- daily.weather.tidy.a %>%
  mutate(month = month(date), 
         year = year(date),
         yday = yday(date)) %>%   
  mutate (prcp.mm = prcp/10, 
          tmax.C = tmax/10, 
          tmin.C = tmin/10, 
          tavg.C = ((tmin.C+tmax.C)/2))

write.csv(daily.weather.tidy.a, "preppedData/daily.weather.csv", row.names = FALSE)



```


```{r import daily Amherst weather data made above}

daily <- read_csv("preppedData/daily.weather.csv",
                  col_types = cols(date = col_date(format = "%Y-%m-%d"))) %>% 
  dplyr::select(-prcp, -tmax, -tmin)

```


```{r calculate total precipitation for 30 days before}

sample.dates <- macro.abio %>% 
  dplyr::select(date) %>% 
  distinct() %>% 
  mutate(month = month(date), 
         year = year(date),
         yday = yday(date)) 



precip <- daily  %>%
  group_by(year) %>% 
  arrange(year, yday) %>% 
  mutate(mon.precip = rollapply(data = prcp.mm,
                                width = 30,
                                FUN = sum,
                                align = "right",
                                fill = NA,
                                na.rm = TRUE))

precip.r <- precip %>% 
  dplyr::select(year, yday, mon.precip)
  

sample.dates <- left_join(sample.dates, precip.r, by=c("year", "yday"))


```


```{r calculate ADD for 30 days before}


ADD <- daily %>%
  group_by(year) %>% 
  arrange(year, yday) %>% 
  mutate(mon.ADD = rollapply(data = tavg.C,
                                width = 30,
                                FUN = sum,
                                align = "right",
                                fill = NA,
                                na.rm = TRUE)) 


ADD.r <- ADD %>% 
  dplyr::select(year, yday, mon.ADD)

sample.weather <- left_join(sample.dates, ADD.r, by=c("year", "yday"))

```

```{r join macro and weather data}


weather <- sample.weather %>% 
  dplyr::select(date, mon.precip, mon.ADD)

macro.abio <- left_join(macro.abio, weather, by="date") 

```


##Import USGS Discharge Data

```{r USGS discharge data}

#Use the dataRetrieval package to download USGS data

#parameter codes
  #00010 = "Wtemp", #water temp in C
  #00045 = "Precip", #precipitation in inches
  #00060 = "Flow", #discharge in cfs
  #00065 = "GH", #gage height in feet
  #00095 = "SpecCond",
  #00300 = "DO",
  #00400 = "pH",
  #62611 = "GWL",
  #63680 = "Turb",
  #72019 = "WLBLS"

#stat codes
  #00001 = max
  #00002 = min
  #00003 = mean
  #00006 = sum
  #00007 = mode
  #00008 = median


#USGS 01171500 MILL RIVER AT NORTHAMPTON, MA
avail.data <- whatNWISdata(siteNumber = '01171500')



#check the information about the parameter
parameterINFO <- readNWISpCode("00060")
colnames(parameterINFO)
parameterINFO$parameter_nm


#Raw daily data:
daily.disch <- readNWISdv(siteNumber = "01171500",
                           parameterCd = "00060", 
                           startDate = "2018-01-01", 
                           endDate = "2019-12-31",
                           statCd = "00003")

attr(daily.disch, "siteInfo")



daily.disch.df <- daily.disch %>% 
  dplyr::select(Date, discharge = X_00060_00003) %>% 
  mutate(year = year(Date),
         month = month(Date),
         yday = yday(Date))



#Raw instantaneous data:
raw.disch <- readNWISdata(sites = "01171500",
                            service = "iv",
                           parameterCd = "00060", 
                           startDate = "2018-01-01", 
                           endDate = "2019-12-31")


daily.disch2 <- raw.disch %>% 
  dplyr::select(dateTime, discharge = X_00060_00000) %>% 
  mutate(date = date(dateTime),
          year = year(dateTime),
         month = month(dateTime),
         yday = yday(dateTime)) %>% 
  group_by(date, year, yday) %>% 
  dplyr::summarize(mean.discharge = mean(discharge, na.rm = TRUE),
                   median.discharge = median(discharge, na.rm = TRUE))


```


```{r USGS gage height data}

#Use the dataRetrieval package to download USGS data

#parameter codes
  #00010 = "Wtemp", #water temp in C
  #00045 = "Precip", #precipitation in inches
  #00060 = "Flow", #discharge in cfs
  #00065 = "GH", #gage height in feet
  #00095 = "SpecCond",
  #00300 = "DO",
  #00400 = "pH",
  #62611 = "GWL",
  #63680 = "Turb",
  #72019 = "WLBLS"



#check the information about the parameter
parameterINFO <- readNWISpCode("00065")
colnames(parameterINFO)
parameterINFO$parameter_nm

#Raw instantaneous data:
raw.gh <- readNWISdata(sites = "01171500",
                            service = "iv",
                           parameterCd = "00065", 
                           startDate = "2018-01-01", 
                           endDate = "2019-12-31")


daily.gh <- raw.gh %>% 
  dplyr::select(dateTime, gh = X_00065_00000) %>% 
  mutate(date = date(dateTime),
          year = year(dateTime),
         month = month(dateTime),
         yday = yday(dateTime)) %>% 
  group_by(date, year, yday) %>% 
  dplyr::summarize(mean.gh = mean(gh, na.rm = TRUE),
                   median.gh = median(gh, na.rm = TRUE))



```


```{r join discharge and gage height data}

#note discharge is in cfs and gage height is in feet

usgs.df <- daily.disch2 %>% 
  dplyr::select(date, mean.discharge, median.discharge) %>% 
  left_join(., daily.gh %>% 
              dplyr::select(date, mean.gh, median.gh)) 
            

write.csv(usgs.df, "preppedData/mill.usgs.csv", 
          row.names = FALSE)


```

```{r import daily Amherst weather data made above}

usgs.df <- read_csv("preppedData/mill.usgs.csv",
                  col_types = cols(date = col_date(format = "%Y-%m-%d"))) 

```


```{r join macro and USGS data}

#make year a factor so that the data frames will join
usgs.df$year <- as.factor(usgs.df$year)

#join the macro and USGS data together
macro.abio <- left_join(macro.abio, usgs.df)


# save the fully merged and prepared data file
# For the metadata describing everything in the mill.macro.csv file see mill_macro_metadata.txt text file included in the Project folder 

write.csv(macro.abio, "preppedData/mill.macro.csv", 
          row.names = FALSE)

```

