---
title: "NEON CARI Data"
author: "Marney Pratt"
date: "10/15/2020"
output: html_notebook
---

# Load Packages  

```{r setup, include=FALSE}

# load libraries 
library(tidyverse) # for readr (import data) and dplyr (manipulate data)
library(lubridate) #for dealing with dates and times
library(neonUtilities) #to download NEON data
library(padr) # time-series data preparation functions
library(zoo) ## for dealing with time series

```

# Downloading NEON Macroinvertebrate Data  

See the NEON website for more information on the [Macroinvertebrate collection Data Product](https://data.neonscience.org/data-products/DP1.20120.001) 

```{r download all NEON macroinvertebrate data, eval=FALSE, include=FALSE}

###I ran this code once and saved the output, then I turned off running this chunk

# Macroinvert Data Product ID
my_dpid <- 'DP1.20120.001'


# get all tables for these sites from the NEON API 
# Remove token=Sys.getenv("NEON_TOKEN") line if you do not have your own API token for NEON
all_tabs_inv <- neonUtilities::loadByProduct(
  dpID = my_dpid,
  site = 'CARI',
  package = 'expanded',
  check.size = FALSE,
  token=Sys.getenv("NEON_TOKEN"))

# extract items from list and put in R env. 
all_tabs_inv %>% list2env(.GlobalEnv)


#save readme tibble as csv file
write.csv(readme_20120, "NEONinfo/NEON_metadata.csv")

#save file with code definitions
write.csv(categoricalCodes_20120, "NEONinfo/categoricalCodes.csv")


# extract location data into a separate table
CARI_fieldData <- inv_fieldData %>%
  
# extract year and date from collectDate, add as new column
  mutate(
    year = year(collectDate), 
    date = date(collectDate)) %>% 
  
  # keep only the columns listed below
  select(siteID, 
         domainID,
         namedLocation, 
         decimalLatitude, 
         decimalLongitude, 
         elevation,
         sampleID, 
         eventID, 
         date,
         year, 
         habitatType, 
         samplerType,
         benthicArea) %>%
  
  # keep rows with unique combinations of values, 
  # i.e., no duplicate records
  distinct()

#Save this data file
write.csv(CARI_fieldData, "NEONdata/CARI/CARI_fieldData.csv")


# Make the observation table.
# start with inv_taxonomyProcessed
CARI.macro <- inv_taxonomyProcessed %>% 
  
  # select a subset of columns from
  # inv_taxonomyProcessed
  select(sampleID,
         domainID,
         siteID,
         collectDate,
         estimatedTotalCount,
         acceptedTaxonID,
         phylum, subphylum, class, subclass, order, family, subfamily, tribe, genus, 
         scientificName,
         taxonRank) %>%
  
  # Join the columns selected above with two 
  # columns from inv_fieldData (the two columns 
  # are sampleID and benthicArea)
  left_join(CARI_fieldData %>% 
              select(sampleID, eventID, date, year, 
                     habitatType, samplerType,
                     benthicArea)) %>%

  # some new columns called 'variable_name', 
  # 'value', and 'unit', and assign values for 
  # all rows in the table.
  # variable_name and unit are both assigned the 
  # same text string for all rows. 
  mutate(inv_dens = estimatedTotalCount / benthicArea,
         inv_dens_unit = 'count per square meter') 


#import master taxa list with tolerance values and FFGs 
neon.mill.taxa <- read_csv("data/neon.mill.taxa.csv")

#select needed variables
neon.mill.taxa2 <- neon.mill.taxa %>%  
  dplyr::select(acceptedTaxonID, organism_aggr, tolerance, FFG, FFG2)

#merge NEON macro data with master taxa data
CARI.macro <- left_join(CARI.macro, neon.mill.taxa2, by="acceptedTaxonID")


# save macro data file for this site
write.csv(CARI.macro, "NEONdata/CARI/CARI.macro.csv", 
          row.names = FALSE)



```





# Downloading NEON Periphyton, seston, and phytoplankton collection Data  

See the NEON website for more information on the [Periphyton, seston, and phytoplankton collection Data Product](https://data.neonscience.org/data-products/DP1.20166.001) 

```{r download all NEON periphyton, seston data, eval=FALSE, include=FALSE}

###I ran this code once and saved the output, then I turned off running this chunk

# Periphyton, seston, and phytoplankton collection Data Product ID
my_dpid <- 'DP1.20166.001'


# get all tables for these sites from the NEON API 
# Remove token=Sys.getenv("NEON_TOKEN") line if you do not have your own API token for NEON
all_tabs_alg <- neonUtilities::loadByProduct(
  dpID = my_dpid,
  site = 'CARI',
  package = 'expanded',
  check.size = FALSE,
  token=Sys.getenv("NEON_TOKEN"))

# extract items from list and put in R env. 
all_tabs_alg %>% list2env(.GlobalEnv)


#save readme tibble as csv file
write.csv(readme_20166, "NEONinfo/NEON_metadata.csv")

#save file with code definitions
write.csv(categoricalCodes_20166, "NEONinfo/categoricalCodes.csv")


# extract location data into a separate table
CARI_alg_fieldData <- alg_fieldData %>%
  
# extract year and date from collectDate, add as new column
  mutate(
    year = year(collectDate), 
    date = date(collectDate)) %>% 
  
  # keep only the columns listed below
  dplyr::select(date,
                year,
                domainID,
                siteID, 
                parentSampleID, 
                eventID,
                namedLocation, 
                decimalLatitude, 
                decimalLongitude, 
                elevation,
                habitatType, 
                algalSampleType,
                samplerType,
                benthicArea,
                substratumSizeClass) %>% 
  dplyr::rename(., algSampleID = parentSampleID) %>% 
  
  # keep rows with unique combinations of values, 
  # i.e., no duplicate records
  distinct()

#Save this data file
write.csv(CARI_alg_fieldData, "NEONdata/CARI/CARI_alg_fieldData.csv")


# Make the observation table.
# start with alg_taxonomyProcessed
CARI.alg <- alg_taxonomyProcessed %>% 
  
  dplyr::rename(., algSampleID = sampleID) 

CARI.alg <- CARI.alg %>% 
  
  # select a subset of columns from
  # alg_taxonomyProcessed
  dplyr::select(domainID,
         siteID,
         namedLocation,
         collectDate,
         algSampleID,
         perBottleSampleVolume,
         algalParameter,
         algalParameterValue,
         acceptedTaxonID,
         division, class, order, family, genus, 
         specificEpithet, infraspecificEpithet,
         scientificName,
         taxonRank) %>%
  
  # Join the columns selected above with  
  # columns from alg_fieldData 
  left_join(CARI_alg_fieldData %>% 
              dplyr::select(algSampleID, eventID, date, year, 
                     habitatType, algalSampleType,samplerType,
                     benthicArea, substratumSizeClass)) %>%

  # some new columns called 'variable_name', 
  # 'value', and 'unit', and assign values for 
  # all rows in the table.
  # variable_name and unit are both assigned the 
  # same text string for all rows. 
  mutate(alg_dens = algalParameterValue / benthicArea,
         alg_dens_unit = 'count or biovolume per square meter') 


# make master algal taxa list
CARI.alg.taxa <- CARI.alg %>% 
  dplyr::select(9:18) %>% 
  distinct()
  
# save master algal taxa list
write.csv(CARI.alg.taxa, "NEONdata/CARI/CARI.alg.taxa.csv", 
          row.names = FALSE)

# remove taxonomy to simplify data file
CARI.alg <- CARI.alg %>% 
  dplyr::select(-(9:18))

# save algae data file for this site
write.csv(CARI.alg, "NEONdata/CARI/CARI.alg.csv", 
          row.names = FALSE)



```



# Download Aquatic Instrument Data  

This webpage is helpful for downloading Aquatic Instrument System Data:
[Download and work with NEON Aquatic Instrument Data](https://www.neonscience.org/explore-neon-ais-data)


### product IDs
  water quality: DP1.20288.001
  air temperature: DP1.00002.001
  relative humidity: DP1.00098.001
  PAR at water surface:  DP1.20042.001
  Precipitation: DP1.00006.001
  Gage height: DP1.20267.001  
  
  
  
  
## Download Water Quality Data Product  

See the NEON website for more information on the [Water Quality Data Product](https://data.neonscience.org/data-products/DP1.20288.001)  


```{r download NEON water quality data, eval=FALSE, include=FALSE}

###I ran this code once and saved the output, then I turned off running this chunk

# Water quality Data Product ID
my_dpid <- 'DP1.20288.001'


# download data from the NEON API 
# Remove token=Sys.getenv("NEON_TOKEN") line if you do not have your own API token for NEON
waq <- neonUtilities::loadByProduct(
  dpID = my_dpid,
  site = "CARI",
  package = 'expanded',
  check.size = FALSE,
  token=Sys.getenv("NEON_TOKEN"))

# extract items from list and put in R env. 
waq %>% list2env(.GlobalEnv)

write.csv(variables_20288, "NEONinfo/waq.vars.csv", 
          row.names = FALSE)

write.csv(readme_20288, "NEONinfo/waq.readme.csv", 
          row.names = FALSE)


waq_sum <- waq_instantaneous %>%
                dplyr::select(1:6,
                              specificConductance,
                              dissolvedOxygen,
                              dissolvedOxygenSaturation,
                              pH,
                              chlorophyll,
                              turbidity,
                              fDOM) %>%
                dplyr::mutate(date = date(startDateTime)) %>%
                group_by(domainID, siteID, horizontalPosition, date) %>%
                dplyr::summarize(
                      mean.SpCond = mean(specificConductance, na.rm = TRUE),
                      median.SpCond = median(specificConductance, na.rm = TRUE),
                      mean.DO = mean(dissolvedOxygen, na.rm = TRUE),
                      median.DO = median(dissolvedOxygen, na.rm = TRUE),
                      mean.DOsat = mean(dissolvedOxygenSaturation, na.rm = TRUE),
                      median.DOsat = median(dissolvedOxygenSaturation, na.rm = TRUE),
                      mean.pH = mean(pH, na.rm = TRUE),
                      median.pH = median(pH, na.rm = TRUE),
                      mean.chl = mean(chlorophyll, na.rm = TRUE),
                      median.chl = median(chlorophyll, na.rm = TRUE),
                      mean.turb = mean(turbidity, na.rm = TRUE),
                      median.turb = median(turbidity, na.rm = TRUE),
                      mean.fDOM = mean(fDOM, na.rm = TRUE),
                      median.fDOM = median(fDOM, na.rm = TRUE)
                                  )

write.csv(waq_sum, "NEONdata/CARI/CARI.waq.csv", 
          row.names = FALSE)

```




## Download the Photosynthetically active radiation at water surface Data Product

See the NEON website for more information on the [Photosynthetically active radiation at water surface Data Product](https://data.neonscience.org/data-products/DP1.20042.001)



```{r download NEON PAR data, eval=FALSE, include=FALSE}

###I ran this code once and saved the output, then I turned off running this chunk


# PAR Data Product ID
my_dpid <- 'DP1.20042.001'

# get all tables for these sites from the NEON API
# Remove token=Sys.getenv("NEON_TOKEN") line if you do not have your own API token for NEON
parws <- neonUtilities::loadByProduct(
          dpID = my_dpid,
          site = "CARI",
          package = 'expanded',
          check.size = FALSE,
          token=Sys.getenv("NEON_TOKEN"))


# extract items from list and put in R env.
parws %>% list2env(.GlobalEnv)


write.csv(variables_20042, "NEONinfo/parws.vars.csv",
row.names = FALSE)

write.csv(readme_20042, "NEONinfo/parws.readme.csv",
row.names = FALSE)



#calculate the average PAR for each date
parws_daily <- PARWS_1min %>%
                dplyr::select(1:7) %>%
                dplyr::mutate(date = date(startDateTime)) %>%
                group_by(domainID, siteID, horizontalPosition, date) %>%
                dplyr::summarize(
                      mean.parws = mean(PARMean, na.rm = TRUE),
                      median.parws = median(PARMean, na.rm = TRUE),
                                  )

write.csv(parws_daily, "NEONdata/CARI/CARI.parws.csv", 
          row.names = FALSE)

```


# Import Data  

After running the above chunks once to download all the data, then just import all the resulting data files running the code below  

```{r Import prepped CARI.macro.csv file, include=FALSE}

# this will import the NEON macro data for this one site
CARI.macro <- read_csv("NEONdata/CARI/CARI.macro.csv")

```

```{r Import prepped CARI.alg.csv file, include=FALSE}

# this will import the NEON Periphyton, seston, and phytoplankton data for this one site
CARI.alg <- read_csv("NEONdata/CARI/CARI.alg.csv")

```


```{r import and join environmental variables, include=FALSE}


#import PAR at water surface data
CARI.parws <- read_csv("NEONdata/CARI/CARI.parws.csv")



```


## Join NEON Macroinvertebrate and Environmental Variables

```{r join NEON macros and environmental vars for CARI}


#filter for upstream sensor sets (note - no fDOM)
CARI.parws.up <- CARI.parws %>% 
        dplyr::filter(horizontalPosition == 101 | horizontalPosition == 111) %>% 
        group_by(domainID, siteID, date) %>% 
        dplyr::summarize(
                      mean.parws = mean(mean.parws, na.rm = TRUE),
                      median.parws = mean(median.parws, na.rm = TRUE))

#filter for downstream sensor sets 
CARI.parws.down <- CARI.parws %>% 
        dplyr::filter(horizontalPosition == 102 | horizontalPosition == 112) %>% 
        group_by(domainID, siteID, date) %>% 
        dplyr::summarize(
                      mean.parws = mean(mean.parws, na.rm = TRUE),
                      median.parws = mean(median.parws, na.rm = TRUE))



# join upstream sensor set data to macro data
CARI.macro.up <- left_join(CARI.macro, CARI.parws.up)

# save the fully merged and prepared data file
# For the metadata describing everything in the file see _____ text file included in the Project folder 

write.csv(CARI.macro.up, "preppedData/CARI.macro.up.csv", 
          row.names = FALSE)
  

# join downstream sensor set data to macro data
CARI.macro.down <- left_join(CARI.macro, CARI.parws.down) 


# save the fully merged and prepared data file
# For the metadata describing everything in the file see _____ text file included in the Project folder 

write.csv(CARI.macro.down, "preppedData/CARI.macro.down.csv", 
          row.names = FALSE)


```

## Join NEON Periphyton and Seston Data and Environmental Variables

```{r join NEON algae and environmental vars for CARI}


#filter for upstream sensor sets (note - no fDOM)
CARI.parws.up <- CARI.parws %>% 
        dplyr::filter(horizontalPosition == 101 | horizontalPosition == 111) %>% 
        group_by(domainID, siteID, date) %>% 
        dplyr::summarize(
                      mean.parws = mean(mean.parws, na.rm = TRUE),
                      median.parws = mean(median.parws, na.rm = TRUE))

#filter for downstream sensor sets 
CARI.parws.down <- CARI.parws %>% 
        dplyr::filter(horizontalPosition == 102 | horizontalPosition == 112) %>% 
        group_by(domainID, siteID, date) %>% 
        dplyr::summarize(
                      mean.parws = mean(mean.parws, na.rm = TRUE),
                      median.parws = mean(median.parws, na.rm = TRUE))



# join upstream sensor set data to algae data
CARI.alg.up <- left_join(CARI.alg, CARI.parws.up)

# save the fully merged and prepared data file
write.csv(CARI.alg.up, "preppedData/CARI.alg.up.csv", 
          row.names = FALSE)
  

# join downstream sensor set data to algae data
CARI.alg.down <- left_join(CARI.alg, CARI.parws.down) 


# save the fully merged and prepared data file
write.csv(CARI.alg.down, "preppedData/CARI.alg.down.csv", 
          row.names = FALSE)


```


