---
title: "NEON WLOU Data"
author: "Marney Pratt"
date: "10/25/2020"
output: html_notebook
---

# Load Packages  

```{r setup, include=FALSE}

# load libraries 
library(tidyverse) # for readr (import data) and dplyr (manipulate data)
library(lubridate) #for dealing with dates and times
library(neonUtilities) #to download NEON data
library(padr) # time-series data preparation functions
library(zoo) ## for dealing with time series

```

# Downloading NEON Macroinvertebrate Data  

```{r download all NEON macroinvertebrate data, eval=FALSE, include=FALSE}

###I ran this code once and saved the output, then I turned off running this chunk

# Macroinvert Data Product ID
my_dpid <- 'DP1.20120.001'


# get all tables for these sites from the NEON API 
# Remove token=Sys.getenv("NEON_TOKEN") line if you do not have your own API token for NEON
all_tabs_inv <- neonUtilities::loadByProduct(
  dpID = my_dpid,
  site = 'WLOU',
  package = 'expanded',
  check.size = FALSE,
  token=Sys.getenv("NEON_TOKEN"))

# extract items from list and put in R env. 
all_tabs_inv %>% list2env(.GlobalEnv)


#save readme tibble as csv file
write.csv(readme_20120, "NEONinfo/NEON_metadata.csv")

#save file with code definitions
write.csv(categoricalCodes_20120, "NEONinfo/categoricalCodes.csv")


# extract location data into a separate table
WLOU_fieldData <- inv_fieldData %>%
  
# extract year and date from collectDate, add as new column
  mutate(
    year = year(collectDate), 
    date = date(collectDate)) %>% 
  
  # keep only the columns listed below
  select(siteID, 
         domainID,
         namedLocation, 
         decimalLatitude, 
         decimalLongitude, 
         elevation,
         sampleID, 
         eventID, 
         date,
         year, 
         habitatType, 
         samplerType,
         benthicArea) %>%
  
  # keep rows with unique combinations of values, 
  # i.e., no duplicate records
  distinct()

#Save this data file
write.csv(WLOU_fieldData, "NEONdata/WLOU/WLOU_fieldData.csv")


# Make the observation table.
# start with inv_taxonomyProcessed
WLOU.macro <- inv_taxonomyProcessed %>% 
  
  # select a subset of columns from
  # inv_taxonomyProcessed
  select(sampleID,
         domainID,
         siteID,
         collectDate,
         estimatedTotalCount,
         acceptedTaxonID,
         phylum, subphylum, class, subclass, order, family, subfamily, tribe, genus, 
         scientificName,
         taxonRank) %>%
  
  # Join the columns selected above with two 
  # columns from inv_fieldData (the two columns 
  # are sampleID and benthicArea)
  left_join(WLOU_fieldData %>% 
              select(sampleID, eventID, date, year, 
                     habitatType, samplerType,
                     benthicArea)) %>%

  # some new columns called 'variable_name', 
  # 'value', and 'unit', and assign values for 
  # all rows in the table.
  # variable_name and unit are both assigned the 
  # same text string for all rows. 
  mutate(inv_dens = estimatedTotalCount / benthicArea,
         inv_dens_unit = 'count per square meter') 


#import master taxa list with tolerance values and FFGs 
neon.mill.taxa <- read_csv("data/neon.mill.taxa.csv")

#select needed variables
neon.mill.taxa2 <- neon.mill.taxa %>%  
  dplyr::select(acceptedTaxonID, organism_aggr, tolerance, FFG, FFG2)

#merge NEON macro data with master taxa data
WLOU.macro <- left_join(WLOU.macro, neon.mill.taxa2, by="acceptedTaxonID")


# save macro data file for this site
write.csv(WLOU.macro, "NEONdata/WLOU/WLOU.macro.csv", 
          row.names = FALSE)



```




## Import NEON Macroinvertebrate Data for WLOU Site

```{r Import prepped neon.macro.csv file, include=FALSE}

# this file is the result of the above chunk
# run the chuk above once to create this file
# then after that, don't run the above chunk and only run this one 
# this will import the NEON macro data for this one site
WLOU.macro <- read_csv("NEONdata/WLOU/WLOU.macro.csv")

```




# Download Aquatic Instrument Data  

This webpage is helpful for downloading Aquatic Instrument System Data:
[Download and work with NEON Aquatic Instrument Data](https://www.neonscience.org/explore-neon-ais-data)


### product IDs
  water quality: DP1.20288.001
  air temperature: DP1.00002.001
  relative humidity: DP1.00098.001
  PAR at water surface:  DP1.20042.001
  Precipitation: DP1.00006.001
  Gage height: DP1.20267.001  
  
  
## Download the Precipitation Data Product  

See the NEON website for more information on the [Precipitation Data Product](https://data.neonscience.org/data-products/DP1.00006.001)  


```{r download NEON precipitation data, eval=FALSE, include=FALSE}

###I ran this code once and saved the output, then I turned off running this chunk


# Precipitation Data Product ID
my_dpid <- 'DP1.00006.001'


# download data from the NEON API 
# Remove token=Sys.getenv("NEON_TOKEN") line if you do not have your own API token for NEON
precip <- neonUtilities::loadByProduct(
  dpID = my_dpid,
  # many aquatic sites don't have precipitation sensors,
  # pick the nearest terrestrial site
  site = "WLOU",
  package = 'expanded',
  check.size = FALSE,
  token=Sys.getenv("NEON_TOKEN"))

# extract items from list and put in R env. 
precip %>% list2env(.GlobalEnv)


write.csv(variables_00006, "NEONinfo/precip.vars.csv",
row.names = FALSE)

write.csv(readme_00006, "NEONinfo/precip.readme.csv",
row.names = FALSE)



#calculate the total precipitation for each date
precip_daily <- SECPRE_30min %>%
                dplyr::select(1:6,
                    secPrecipBulk) %>%
                    dplyr::mutate(date = date(startDateTime),
                    year = year(startDateTime),
                    yday = yday(startDateTime)) %>%
                    group_by(domainID, siteID, date, year, yday) %>%
                    dplyr::summarize(
                    daily.precip = sum(secPrecipBulk, na.rm = TRUE)
                    )

# calculate the total monthly precipitation 
# by summing the daily precipitation for 30 days before each date

precip <- precip_daily %>%
                group_by(year) %>%
                arrange(year, yday) %>%
                mutate(mon.precip = rollapply(data = daily.precip,
                width = 30,
                FUN = sum,
                align = "right",
                fill = NA,
                na.rm = TRUE))



write.csv(precip, "NEONdata/WLOU/WLOU.precip.csv",
row.names = FALSE)


```




## Import Environmental Variables  

After running the above chunks once to download all the environmental variables, then just import all the resulting data files running the code below  

```{r import and join environmental variables, include=FALSE}


#import precipitation data
WLOU.precip<- read_csv("NEONdata/WLOU/WLOU.precip.csv")



```


## Join NEON Macroinvertebrate Data and Environmental Variables

```{r join NEON macros and environmental vars for WLOU}


# select columns needed
WLOU.precip.r <- WLOU.precip %>% dplyr::select(date, mon.precip)

# join precip to macro data
WLOU.macro.precip <- left_join(WLOU.macro, WLOU.precip.r, by = "date") 



# save the fully merged and prepared data file
# For the metadata describing everything in the file see _____ text file included in the Project folder 

write.csv(WLOU.macro.precip, "preppedData/WLOU.macro.precip.csv", 
          row.names = FALSE)


```



